{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2017-10-27 sftrends_manu.tex
minor edits to maybe final draft

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Fri Oct 20 11:14:05 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\section*{Abstract}
\noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.  Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.  Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).  Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from 433 at D19 to 571 at C3).  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10}. Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods. Flow-normalization predictions for each month of each year were based on the average of predictions for flow values that occur in the same month across all years, weighted within each specific month and year for every observation. Flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. Annual groupings were chosen as approximate twenty year midpoints in the time series and seasonal groupings were chosen to evaluate inter-annual changes while keeping season constant.

@@

A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), thirteen comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.  Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}). The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

@@

The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, clam biomass was associated with a decrease in \ac{chla} concentration, as shown by others \citep{Alpine92,Thompson08}. Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.  Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.  Flow effects on phytoplankton production have also changed over time. In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the estuary (\cref{fig:clmchl}j). Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

@@

As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time. \ac{wrtds} models can facilitate descriptions of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.  We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.  The results from \ac{wrtds} are also consistent with described trends of nutrient loads from point sources (e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}) demonstrating that these changes are not unexpected.  Consequently, we are not detracting from the potential implications of such increases. The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels. Therefore, methods that adjust for the effects of these factors are critical when studying long-term records to assess the impacts or effectiveness of load increases or management actions, respectively.

Combined with additional data, our results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics. Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.  For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow effects.  Alternative methods could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).  Overall, quantitative interpretations of multiple factors can provide a basis for identifying links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.

@@

We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program. We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Elizabeth George for providing comments on an earlier draft. This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory. The authors declare no competing financial interest. The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.

@@

\caption{Model predictions for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \cref{fig:delt_map}).  Results are annually-averaged for each water year from October to September.  Points are model predictions and lines are flow-normalized predictions.}\label{fig:prddat}
\end{figure}

@@

% site descrip
%latex.default(totab, file = "", rowlabel = "Station", caption = cap.val,     caption.loc = "top", rgroup = rgrps, n.rgroup = c(4, 3, 3),     cgroup = cgrps, n.cgroup = c(1, 1, 2, 2, 2), rowname = stats,     label = "tab:stdescrp")%
\begin{table}[!tbp]
\caption{Monitoring stations in the upper San Francico Estuary used to evaluate nitrogen trends. Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station. Median values are reported in mg L$^{-1}$.\label{tab:stdescrp}} 
\begin{center}
\begin{tabular}{lrcrcrrcrrcrr}
\hline\hline
\multicolumn{1}{l}{\bfseries Station}&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}\tabularnewline
\cline{6-7} \cline{9-10} \cline{12-13}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Lat}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Lon}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}\tabularnewline
\hline
{\bfseries Peripheral}&&&&&&&&&&&&\tabularnewline
~~C3&$38.35$&&$-121.55$&&$569$&$0.36$&&$569$&$0.22$&&$571$&$0.13$\tabularnewline
~~C10&$37.68$&&$-121.27$&&$539$&$1.44$&&$539$&$0.04$&&$558$&$1.40$\tabularnewline
~~MD10&$38.04$&&$-121.42$&&$548$&$0.31$&&$548$&$0.03$&&$570$&$0.28$\tabularnewline
~~P8&$37.98$&&$-121.38$&&$556$&$1.46$&&$556$&$0.12$&&$563$&$1.20$\tabularnewline
\hline
{\bfseries Interior}&&&&&&&&&&&&\tabularnewline
~~D19&$38.04$&&$-121.61$&&$433$&$0.35$&&$435$&$0.04$&&$462$&$0.31$\tabularnewline
~~D26&$38.08$&&$-121.57$&&$556$&$0.38$&&$556$&$0.09$&&$565$&$0.29$\tabularnewline
~~D28&$37.97$&&$-121.57$&&$529$&$0.38$&&$535$&$0.03$&&$555$&$0.33$\tabularnewline
\hline
{\bfseries Suisun}&&&&&&&&&&&&\tabularnewline
~~D4&$38.06$&&$-121.82$&&$546$&$0.38$&&$546$&$0.05$&&$565$&$0.32$\tabularnewline
~~D6&$38.04$&&$-122.12$&&$534$&$0.45$&&$534$&$0.08$&&$562$&$0.35$\tabularnewline
~~D7&$38.12$&&$-122.04$&&$535$&$0.44$&&$535$&$0.06$&&$561$&$0.36$\tabularnewline
\hline
\end{tabular}\end{center}
\end{table}

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Fri Oct 20 11:24:35 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\section*{Abstract}
\noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.  Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.  Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed. Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although a consistent decline began in the mid-1990s and early 2000s. We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \acl{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \acl{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; \textit{interior} Delta stations D19, D26, D28; and \textit{Suisun} stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal ($\beta_3$, $\beta_4$) trend, and $Q$ is the flow variable (either flow or salinity depending on station).

The \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from 433 at D19 to 571 at C3).  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10} or tidal variation \citep{Beck15}. Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods. Flow-normalization predictions for each month of each year were based on the average of predictions for flow values that occur in the same month across all years, weighted within each specific month and year for every observation. Flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. Annual groupings were chosen as approximate twenty year midpoints in the time series and seasonal groupings were chosen to evaluate inter-annual changes while keeping season constant.

@@

A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), thirteen comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.  Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}). The most notable change in the flow-normalized results was an overall decrease (less positive trend or greater negative trend) in concentrations for most sites in the summer and fall seasons for 1996-2013.  More statistically significant trends were also observed with the flow-normalized results.

@@

The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, clam biomass was associated with a decrease in \ac{chla} concentration, as shown by others \citep{Alpine92,Thompson08}. Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.  Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.  Flow effects on phytoplankton production have also changed over time. In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the estuary (\cref{fig:clmchl}j). Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanisms demonstrated by \ac{wrtds} provide additonal quantitative evidence of factors that drive water quality in the Delta.

@@

As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time. \ac{wrtds} models can facilitate descriptions of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.  We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.  The results from \ac{wrtds} are also consistent with described trends of nutrient loads from point sources \citep[e.g., Sacramento \ac{wwtp} increases and exports to Suisun Bay,][]{Jassby08,Novick14}, demonstrating that these changes are not unexpected.  Consequently, we are not detracting from the potential implications of such increases. The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels. Therefore, methods that adjust for the effects of these factors are critical when studying long-term records to assess the impacts or effectiveness of load increases or management actions, respectively.

Combined with additional data, our results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics. Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.  For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow effects.  Alternative methods could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables \citep[e.g., generalized additive models,][]{Beck17}.  Overall, quantitative interpretations of multiple factors can provide a more comprehensive understanding of relationships between nutrients and primary production, including adverse effects on ecosystem condition.

@@

We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program. We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Susan Elizabeth George for providing comments on an earlier draft. This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory. The authors declare no competing financial interest. The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.

@@

\caption{Model results (points) and flow-normalized predictions (lines) for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \cref{fig:delt_map}).  Results are annually-averaged for each water year from October to September.}\label{fig:prddat}
\end{figure}

@@

% site descrip
%latex.default(totab, file = "", rowlabel = "Station", caption = cap.val,     caption.loc = "top", rgroup = rgrps, n.rgroup = c(4, 3, 3),     cgroup = cgrps, n.cgroup = c(1, 1, 2, 2, 2), rowname = stats,     label = "tab:stdescrp")%
\begin{table}[!tbp]
\caption{Monitoring stations in the upper San Francico Estuary used to evaluate nitrogen trends. Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station. Median values (mg L$^{-1}$) are reported for the entire period of record.\label{tab:stdescrp}} 
\begin{center}
\begin{tabular}{lrcrcrrcrrcrr}
\hline\hline
\multicolumn{1}{l}{\bfseries Station}&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}\tabularnewline
\cline{6-7} \cline{9-10} \cline{12-13}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Lat}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Lon}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}\tabularnewline
\hline
{\bfseries Peripheral}&&&&&&&&&&&&\tabularnewline
~~C3&$38.35$&&$-121.55$&&$569$&$0.36$&&$569$&$0.22$&&$571$&$0.13$\tabularnewline
~~C10&$37.68$&&$-121.27$&&$539$&$1.44$&&$539$&$0.04$&&$558$&$1.40$\tabularnewline
~~MD10&$38.04$&&$-121.42$&&$548$&$0.31$&&$548$&$0.03$&&$570$&$0.28$\tabularnewline
~~P8&$37.98$&&$-121.38$&&$556$&$1.46$&&$556$&$0.12$&&$563$&$1.20$\tabularnewline
\hline
{\bfseries Interior}&&&&&&&&&&&&\tabularnewline
~~D19&$38.04$&&$-121.61$&&$433$&$0.35$&&$435$&$0.04$&&$462$&$0.31$\tabularnewline
~~D26&$38.08$&&$-121.57$&&$556$&$0.38$&&$556$&$0.09$&&$565$&$0.29$\tabularnewline
~~D28&$37.97$&&$-121.57$&&$529$&$0.38$&&$535$&$0.03$&&$555$&$0.33$\tabularnewline
\hline
{\bfseries Suisun}&&&&&&&&&&&&\tabularnewline
~~D4&$38.06$&&$-121.82$&&$546$&$0.38$&&$546$&$0.05$&&$565$&$0.32$\tabularnewline
~~D6&$38.04$&&$-122.12$&&$534$&$0.45$&&$534$&$0.08$&&$562$&$0.35$\tabularnewline
~~D7&$38.12$&&$-122.04$&&$535$&$0.44$&&$535$&$0.06$&&$561$&$0.36$\tabularnewline
\hline
\end{tabular}\end{center}
\end{table}
